Pool Model - A unified Mathematical\\ Modelling Framework for Food Science

### 2025-10-31 paper/pool_model.tex
fix formatting of i.e. and e.g.

\begin{abstract}
    Bacterial consortia are complex systems which are often comprised of multiple distinct species.
    Their dynamics on a population-based level often result from various intra- and inter-species interactions,
    including cooperation or competition.
    Various empirical and mechanistic models have been conceived and applied to problems related to food microbiology.
    In this work we present a generalized framework, which uses multiple distinct "pools" that can distinguish
    between microbial types in different metabolic states or describe the extracellular environment properties.
    It allows us to incorporate the underlying biological processes and construct the \ac{ode} models of
    different complexity levels.
    To illustrate the application scope of the model, multiple different examples of the microbial systems
    are reported.
    Finally, to explore the limitations of the \ac{ode} formulation, we analyze the influence of the spatial effects.
    For this, we construct an \acl{abm} for a two-species system, showcasing that the resulting pool model
    arises naturally from the individual behaviour of many cellular agents.
    We further apply this \ac{abm} to study the role of properties such as diffusion and spatial heterogeneity
    on the dynamics of bacterial growth.
\end{abstract}

@@

In order to achieve a comprehensive description of the target system, it is important to incorporate
all relevant biological phenomena arising in these bacterial systems.
During their reproductive process, bacteria undergo multiple phases~\citep{buchanan_when_1997}, starting in a lag phase which
is necessary to adapt to new environments and initially impedes the overall growth~\citep{rolfe_lag_2012, monod_growth_1949}.
Additionally, bacteria are able to cooperate e.g. via complementary metabolic processes or inhibit
each other e.g. by means of waste production as a metabolic byproduct~\citep{west_social_2007, rainey_evolution_2003, hibbing_bacterial_2010, stubbendieck_bacterial_2016}.
A limited supply of nutrients will also affect the overall reproductive capacity of the colony~\citep{monod_growth_1949}.
Bacteria have thus devised various survival strategies to ensure sufficient access to nutrients.
Effects on a single-cell level such as adhesion~\citep{htuson_bacteriasurface_2013}, migration~\cite{decoene_microscopic_2011} and chemotaxis can lead to complex
phenomena on a population-based level which can aid in the goal of survival.
Cell signaling e.g. in the form of quorum-sensing can lead to collective emergent phenomena such as
biofilm formation which play an important role for many mature microbial communities~\citep{Jin2020,ng_bacterial_2009}.

% \paragraph{ODE models (predictive microbiology, existing, overview, most popular)}
% Existing models in Predictive Microbiology (Primary Models)~\citep{perez-rodriguez_predictive_2012}:
% \begin{itemize}
%     \item An overview of primary models used in food science can be found in~\citep{van_boekel_kinetic_2008}.
%     \item Baranyi-Roberts Model~\citep{baranyi_modeling_1993, baranyi_dynamic_1994}
%     \item modified Gompertz~\citep{Gompertz1825-wi,Gibson1987}
%     \item Buchanan et al.: three-phase linear model for lag phase, exponential growth phase, and stationary phase~\citep{buchanan_when_1997}
%     \item McKellar two-compartment model with bacteria in growing and non-growing state~\citep{mckellar_heterogeneous_1997}
%     \item Inactivation models
% \end{itemize}
% %
% Secondary models:
% \begin{itemize}
%     \item Polynomial models 
%     \item Square Root-Type Models, e.g., Ratkowski for temperature~\citep{ratkowsky_relationship_1982}
%     \item The Cardinal Parameter Model~\citep{zwietering_decision_1992}
% \end{itemize}

@@

In this work, we introduce a new \ac{ode}-based modeling framework which describes different
bacterial species, their metabolic states and external quantities such as nutrients as individual
pools.
The dynamics of bacterial growth can be understood in terms of transitions between the different
bacterial states i.e. pools.
Our model also allows us to capture interactions between these bacterial states and the external
microenvironment.
Compared against the aforementioned methods, this approach turns out to be less complex with general
applicability, low number of parameters and greater explainability for the underlying biological
reality.
To illustrate the generality of the model, multiple examples are reported.
We analyze the spatial limitations of an interacting two-species model and show that it can be fully
understood on an individual-based level in terms of single-cell processes.
Furthermore, we assess how spatial heterogeneity~\citep{mckellar_heterogeneous_1997} can impact the shape of the growth curves.

@@

Then $R=N_R/V$ and $[R]=1/V$, where $N_R$ is the number of resource molecules (or any appropriate unit, e.g., mol).

@@

The relevant question here is whether cells in the pool $D$ still contribute to the limitation of the resources by, e.g., occupying space or consuming important nutrients.

@@

However, in the case of time dependent parameters, e.g., for a dependence on a non-static, dynamic temperature, it appears to be not advantageous to use the analytical results for $L$ and $D$ and work instead with all three equations for $L$, $G$, and $D$.

@@

This survival strategy can appear in response to different external stresses, e.g., nutrient deficiency, temperature stress or extreme fluctuations of environment parameters.

@@

Such situation with 2 bacteria species, a shared resource and produced inhibitor was modeled, e.g., by \citet{lenski_coexistence_1986}.

@@

It is straight forward to make the mechanism of inhibition more explicit by, e.g., introducing a toxin pool.

@@

Another mechanism is the motility of bacteria, e.g., the flagellates \textit{E. coli} or

@@

%     \item \citep{decoene_microscopic_2011}: 2D chemotactic self-propelled bacterial motion of the flagellated bacteria (e.g. E. coli, B. subtilis).

@@

%     Can use citation when talk about general processes we are accounting as growth, uptake of nutrient, production of inhibitor, e.g.,:

@@

The presented pool model was used to describe the dynamics of one- and two-species systems containing the main growth stages and interactions between bacteria in the form of cooperation and competition.
Our approach produces results which exhibit similarities to other well-known predictive microbiology models such as the Baranyi-Roberts model while providing better explainability.
The inclusion of additional metabolic processes or larger number of cell-species can be achieved by extending the list of possible states and defining transitions between them.
As an example, the description of the quorum sensing phenomena can be modelled by adding the equation for the released autoinducer molecules as a part of the microenvironment.
These molecules then affect the same-species bacteria growth controlling the size of the population.
The described above approach provides a primary predictive model with transition rates between different pools as parameters.
To include the influence of external factors, i.e. temperature, pH, pressure, etc., one can simply use the secondary models for the transition rates, e.g., the Ratkowsky model for temperature dependence~\citep{ratkowsky_relationship_1982}.\\
By constructing an \ac{abm}, we have exemplified how close the devised model is to the individual biological 
processes that occur within the bacteria.
This mechanistic single-cell level of understanding of our \ac{abm} is a perspective which can not be easily replicated by other spatial models such as \acp{pde}.
We have investigated the impact of spatial effects in the form of varying diffusion parameters and different initial spatial distributions of bacteria.
Our findings reinforce that the \ac{ode} model is a good simplification in the highly-diffusive regime within a mostly homogenous spatial bacterial distribution.
However, deviations to these assumptions can lead to non-negligible effects, influencing the growth of the colony.

@@

% ================================================================================================
\newpage
\beginsupplement
\section{Appendix}
% ================================================================================================
\subsection{General Mathematical Formalism}
\label{ssec:supplement1}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, all discussed in the main part systems  can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
\label{eq:model_ODE_x}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector describing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentration of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.\\
%
The advantage of the following formulation is its generality.
Depending on the knowledge about the system as well as the needed level of accuracy one can define an arbitrary number of states $M$ and control the complexity of the model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.\\
%
To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_i = \sum_k \mbx_i^k.
\label{eq:N_pro_species}
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N = \sum_i \sum_k \mbx_i^k.
\label{eq:N_total}
\end{equation}
%
%
%The three pool system described with equations~\ref{eq:ode_3pools} will be written as
%\begin{equation}
%    \dot \mbx  = A \mbx, \hspace{3mm} A = \begin{pmatrix}
%        -(\mu + \lambda) & 0             & 0      \\
%        \lambda          & \alpha - \mu' & 0      \\
%        \mu              & \mu'          & -\beta 
%    \end{pmatrix}.
%\label{eq:x_repres_3pools}
%\end{equation}
%
Using such matrix representation of the bacterial system, we can rewrite, for example, the three-pool model with limiting resource described with equations~\ref{eq:ode_3pool_resource} in a more compact form.
The concentrations of the pools are then the components of our variable vector $\mbx (t) = (L(t), G(t), D(t))^T$.
%
\begin{align}
    \dot \mbx  &= \mbF(\mbx, \mbm, \mbp, \mbu) = \begin{pmatrix}
        -(\mu + \lambda) & 0               & 0      \\
        \lambda          & \alpha R - \mu' & 0      \\
        \mu              & \mu'            & -\beta 
    \end{pmatrix} \mbx\\
    \dot \mbm &= \mbH (\mbx, \mbm, \mbp, \mbu) = -\frac{\alpha}{N_t} R G.
\end{align}
Separating linear and non-linear parts in two terms yields
\begin{equation}
    \dot \mbx = \begin{pmatrix}
        -(\mu + \lambda) & 0       & 0 \\
         \lambda         & \alpha  & 0 \\
         \mu &  \mu'& - \beta 
    \end{pmatrix} 
    \mbx + \frac{\alpha}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
\label{eq:x_repres_3pools_resource2}
\end{equation}
Here the first term is the linear part and the second term define the quadratic behavior of the system.
%
\paragraph{Generalized Lotka-Volterra}

@@

This means that every contribution does not depend on the position at which it was evaluated, i.e. $[R]_{x=x_c}=[R]$.

>>>>

\begin{abstract}
    Bacterial consortia are complex systems which are often comprised of multiple distinct species.
    Their dynamics on a population-based level often result from various intra- and inter-species interactions,
    including cooperation or competition.
    Various empirical and mechanistic models have been conceived and applied to problems related to food microbiology.
    In this work we present a generalized framework, which uses multiple distinct "pools" that can distinguish
    between microbial types in different metabolic states or describe the extracellular environment properties.
    It allows us to incorporate the underlying biological processes and construct \ac{ode} based models of
    different complexity levels.
    To illustrate the application scope of the model, multiple different examples of microbial systems
    are reported.
    Finally, to explore the limitations of the \ac{ode} formulation, we analyze the influence of spatial effects.
    For this, we construct an \ac{abm} for a two-species system, showcasing that the resulting pool model
    arises naturally from the individual behaviour of many cellular agents.
    We further apply this \ac{abm} to study the role of properties such as diffusion and spatial heterogeneity
    on the dynamics of bacterial growth.
\end{abstract}

@@

In order to achieve a comprehensive description of the target system, it is important to incorporate
all relevant biological phenomena arising in these bacterial systems.
During their reproductive process, bacteria undergo multiple phases~\citep{buchanan_when_1997}, starting in a lag phase which
is necessary to adapt to new environments and initially impedes the overall growth~\citep{rolfe_lag_2012, monod_growth_1949}.
Additionally, bacteria are able to cooperate, e.g., via complementary metabolic processes or inhibit
each other, e.g., by means of waste production as a metabolic byproduct~\citep{west_social_2007, rainey_evolution_2003, hibbing_bacterial_2010, stubbendieck_bacterial_2016}.
A limited supply of nutrients will also affect the overall reproductive capacity of the colony~\citep{monod_growth_1949}.
Bacteria have thus devised various survival strategies to ensure sufficient access to nutrients.
Effects on a single-cell level such as adhesion~\citep{htuson_bacteriasurface_2013}, migration~\cite{decoene_microscopic_2011} and chemotaxis can lead to complex
phenomena on a population-based level which can aid in the goal of survival.
Cell signaling, e.g., in the form of quorum-sensing can lead to collective emergent phenomena such as
biofilm formation which play an important role for many mature microbial communities~\citep{Jin2020,ng_bacterial_2009}.

% \paragraph{ODE models (predictive microbiology, existing, overview, most popular)}
% Existing models in Predictive Microbiology (Primary Models)~\citep{perez-rodriguez_predictive_2012}:
% \begin{itemize}
%     \item An overview of primary models used in food science can be found in~\citep{van_boekel_kinetic_2008}.
%     \item Baranyi-Roberts Model~\citep{baranyi_modeling_1993, baranyi_dynamic_1994}
%     \item modified Gompertz~\citep{Gompertz1825-wi,Gibson1987}
%     \item Buchanan et al.: three-phase linear model for lag phase, exponential growth phase, and stationary phase~\citep{buchanan_when_1997}
%     \item McKellar two-compartment model with bacteria in growing and non-growing state~\citep{mckellar_heterogeneous_1997}
%     \item Inactivation models
% \end{itemize}
% %
% Secondary models:
% \begin{itemize}
%     \item Polynomial models 
%     \item Square Root-Type Models, e.g.,, Ratkowski for temperature~\citep{ratkowsky_relationship_1982}
%     \item The Cardinal Parameter Model~\citep{zwietering_decision_1992}
% \end{itemize}

@@

In this work, we introduce a new \ac{ode}-based modeling framework which describes different
bacterial species, their metabolic states and external quantities such as nutrients as individual
pools.
The dynamics of bacterial growth can be understood in terms of transitions between the different
bacterial states, i.e., pools.
Our model also allows us to capture interactions between these bacterial states and the external
microenvironment.
Compared against the aforementioned methods, this approach turns out to be less complex with general
applicability, low number of parameters and greater explainability for the underlying biological
reality.
To illustrate the generality of the model, multiple examples are reported.
We analyze the spatial limitations of an interacting two-species model and show that it can be fully
understood on an individual-based level in terms of single-cell processes.
Furthermore, we assess how spatial heterogeneity~\citep{mckellar_heterogeneous_1997} can impact the shape of the growth curves.

@@

Then $R=N_R/V$ and $[R]=1/V$, where $N_R$ is the number of resource molecules (or any appropriate unit, e.g.,, mol).

@@

The relevant question here is whether cells in the pool $D$ still contribute to the limitation of the resources by, e.g.,, occupying space or consuming important nutrients.

@@

However, in the case of time dependent parameters, e.g.,, for a dependence on a non-static, dynamic temperature, it appears to be not advantageous to use the analytical results for $L$ and $D$ and work instead with all three equations for $L$, $G$, and $D$.

@@

This survival strategy can appear in response to different external stresses, e.g.,, nutrient deficiency, temperature stress or extreme fluctuations of environment parameters.

@@

Such situation with 2 bacteria species, a shared resource and produced inhibitor was modeled, e.g.,, by \citet{lenski_coexistence_1986}.

@@

It is straight forward to make the mechanism of inhibition more explicit by, e.g.,, introducing a toxin pool.

@@

Another mechanism is the motility of bacteria, e.g.,, the flagellates \textit{E. coli} or

@@

%     \item \citep{decoene_microscopic_2011}: 2D chemotactic self-propelled bacterial motion of the flagellated bacteria (e.g., E. coli, B. subtilis).

@@

%     Can use citation when talk about general processes we are accounting as growth, uptake of nutrient, production of inhibitor, e.g.,,:

@@

The presented pool model was used to describe the dynamics of one- and two-species systems containing the main growth stages and interactions between bacteria in the form of cooperation and competition.
Our approach produces results which exhibit similarities to other well-known predictive microbiology models such as the Baranyi-Roberts model while providing better explainability.
The inclusion of additional metabolic processes or larger number of cell-species can be achieved by extending the list of possible states and defining transitions between them.
As an example, the description of the quorum sensing phenomena can be modelled by adding the equation for the released autoinducer molecules as a part of the microenvironment.
These molecules then affect the same-species bacteria growth controlling the size of the population.
The described above approach provides a primary predictive model with transition rates between different pools as parameters.
To include the influence of external factors, i.e., temperature, pH, pressure, etc., one can simply use the secondary models for the transition rates, e.g.,, the Ratkowsky model for temperature dependence~\citep{ratkowsky_relationship_1982}.\\
By constructing an \ac{abm}, we have exemplified how close the devised model is to the individual biological 
processes that occur within the bacteria.
This mechanistic single-cell level of understanding of our \ac{abm} is a perspective which can not be easily replicated by other spatial models such as \acp{pde}.
We have investigated the impact of spatial effects in the form of varying diffusion parameters and different initial spatial distributions of bacteria.
Our findings reinforce that the \ac{ode} model is a good simplification in the highly-diffusive regime within a mostly homogenous spatial bacterial distribution.
However, deviations to these assumptions can lead to non-negligible effects, influencing the growth of the colony.

@@

% ================================================================================================
\newpage
\beginsupplement
\section{Appendix}
% ================================================================================================
\subsection{General Mathematical Formalism}
\label{ssec:supplement1}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, all discussed in the main part systems  can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
\label{eq:model_ODE_x}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector describing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g.,, temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentration of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.\\
%
The advantage of the following formulation is its generality.
Depending on the knowledge about the system as well as the needed level of accuracy one can define an arbitrary number of states $M$ and control the complexity of the model.
Moreover, opposite to the already known Predictive Microbiology like Baranyi and Roberts one can easily change.\\
%
To calculate the bacterial count for one species $i$, we simply need to sum the tensor/vector $\mbx$ over all possible states:
\begin{equation}
    N_i = \sum_k \mbx_i^k.
\label{eq:N_pro_species}
\end{equation}
While the total bacterial count of all species of our system is
\begin{equation}
    N = \sum_i \sum_k \mbx_i^k.
\label{eq:N_total}
\end{equation}
%
%
%The three pool system described with equations~\ref{eq:ode_3pools} will be written as
%\begin{equation}
%    \dot \mbx  = A \mbx, \hspace{3mm} A = \begin{pmatrix}
%        -(\mu + \lambda) & 0             & 0      \\
%        \lambda          & \alpha - \mu' & 0      \\
%        \mu              & \mu'          & -\beta 
%    \end{pmatrix}.
%\label{eq:x_repres_3pools}
%\end{equation}
%
Using such matrix representation of the bacterial system, we can rewrite, for example, the three-pool model with limiting resource described with equations~\ref{eq:ode_3pool_resource} in a more compact form.
The concentrations of the pools are then the components of our variable vector $\mbx (t) = (L(t), G(t), D(t))^T$.
%
\begin{align}
    \dot \mbx  &= \mbF(\mbx, \mbm, \mbp, \mbu) = \begin{pmatrix}
        -(\mu + \lambda) & 0               & 0      \\
        \lambda          & \alpha R - \mu' & 0      \\
        \mu              & \mu'            & -\beta 
    \end{pmatrix} \mbx\\
    \dot \mbm &= \mbH (\mbx, \mbm, \mbp, \mbu) = -\frac{\alpha}{N_t} R G.
\end{align}
Separating linear and non-linear parts in two terms yields
\begin{equation}
    \dot \mbx = \begin{pmatrix}
        -(\mu + \lambda) & 0       & 0 \\
         \lambda         & \alpha  & 0 \\
         \mu &  \mu'& - \beta 
    \end{pmatrix} 
    \mbx + \frac{\alpha}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
\label{eq:x_repres_3pools_resource2}
\end{equation}
Here the first term is the linear part and the second term define the quadratic behavior of the system.
%
\paragraph{Generalized Lotka-Volterra}

@@

This means that every contribution does not depend on the position at which it was evaluated, i.e., $[R]_{x=x_c}=[R]$.
